\documentclass{article}
\usepackage{iclr2025_conference,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{subcaption}

% \iclrfinalcopy  % commented out for blind submission

\title{Activation Transport: An Optimal-Transport Metric Reveals That\\Cross-Architecture Representations Are Far Less Aligned\\Than Centered Kernel Alignment Suggests}

\author{Anonymous Author \\
Anonymous Institution \\
\texttt{anonymous@example.com} \\
}

\begin{document}
\maketitle

\begin{abstract}
Centered Kernel Alignment (CKA) has become the default tool for comparing internal representations of neural networks, and it routinely reports moderate-to-high similarity between convolutional networks and Vision Transformers trained on the same data. We argue that this conclusion is partly an artifact of how CKA aggregates information, and we propose \emph{Activation Transport} (AT), a complementary similarity metric based on solving an optimal channel-to-channel assignment over per-channel activation distributions. AT computes the 1-Wasserstein distance between the empirical distributions of every pair of channels and then finds the minimum-cost matching by linear-sum assignment; the average matched cost is the AT score (lower means more similar). We compare CNNs (ResNet-18, ResNet-50), Vision Transformers (custom ViT-Small, ViT-B/16, DeiT-B), and MLP-Mixers (custom Mixer, Mixer-B/16) on CIFAR-10, CIFAR-100, and ImageNet-V2.\footnote{Code and data: \url{https://anonymous.4open.science/r/activation-transport}} Across all three benchmarks, cross-architecture AT exceeds same-architecture AT by a factor of $7\times$ on CIFAR-10 (10 seeds; $t=31.27$, $p<10^{-16}$) and $26\times$ on CIFAR-100 (10 seeds; $t=11.00$, $p<10^{-15}$). On ImageNet-V2 we report two complementary ratios: a conservative $\mathbf{1.88\times}$ averaged across three same-architecture baselines (ResNet-50 vs.\ ResNet-50-tv, ViT-B/16 vs.\ DeiT-B, Mixer-B/16 vs.\ Mixer-MIIL), and an anchored $\mathbf{60.5\times}$ (ResNet-50 vs.\ ViT-B/16 against the architecturally cleanest ViT-DeiT baseline; $t=6597$, $p=1.05\times 10^{-13}$). Across all three baselines and four depth fractions, cross-architecture ResNet-vs-ViT AT exceeds same-architecture AT in $\boldsymbol{12/12}$ comparisons, with a worst-case ratio of $1.19\times$ and median $3.0\times$. CKA tells the opposite story: on the same models, cross/same CKA ratios sit around $0.65$--$0.76$, suggesting only moderate dissimilarity. Layer-wise AT grows monotonically with depth in every dataset, and AT consistently clusters attention-based architectures (ViT, Mixer) closer to each other than to convolutional networks. We conclude that AT and CKA measure different aspects of representational geometry, and that a careful choice of the same-architecture baseline is essential for any claim about representational universality.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:introduction}

A central question in deep learning is whether networks with different architectural inductive biases converge to similar internal representations when trained on the same data. The answer has practical implications for transfer learning (can features transfer across architecture families?), model ensembling (do diverse architectures provide complementary representations?), and mechanistic interpretability (do insights about one architecture generalize to others?).

The standard tool for measuring representation similarity is Centered Kernel Alignment \citep[CKA;][]{kornblith2019similarity}, which compares the structure of activation spaces via kernel alignment. CKA has been widely adopted for cross-architecture comparison \citep{raghu2021vision,nguyen2021wide}, typically reporting moderate-to-high similarity between architecturally distinct networks trained on the same task. However, CKA measures global similarity in how networks organize data points relative to each other---it does not assess whether the same \emph{features} (individual neurons or channels) encode the same information.

We introduce \textbf{Activation Transport (AT)}, a metric that directly measures feature-level correspondence by solving an optimal transport problem between channel-wise activation distributions. The key insight is that if two networks learn the same features, there should exist a low-cost assignment between their activation channels---each channel in one network maps to a corresponding channel in the other with a similar activation distribution. AT quantifies the average cost of this optimal assignment.

Our experiments on CIFAR-10 (10 seeds) and CIFAR-100 (10 seeds) reveal four main findings:

\begin{enumerate}
    \item \textbf{Cross-architecture divergence is $\mathbf{7}$--$\mathbf{26\times}$ same-architecture divergence.} On CIFAR-10, ResNet-18 vs.\ ViT-Small yields AT $= 0.250$, while ResNet-18 vs.\ a differently-initialized ResNet-18 yields AT $= 0.028$. On CIFAR-100, this gap widens to $26\times$ (cross mean $= 0.254$, same $= 0.010$; $t = 11.00$, $p = 8.1 \times 10^{-16}$).
    \item \textbf{CKA fails to detect this divergence.} For the same ResNet--ViT comparison on CIFAR-10, CKA reports $0.633$---suggesting moderate similarity and dramatically underestimating the representational gap that AT reveals.
    \item \textbf{Divergence increases with depth.} Layer-wise AT between ResNet and ViT grows from $0.034$ (early layers) to $0.255$ (final representations), demonstrating that architectural inductive biases \emph{compound} rather than converge through the network.
    \item \textbf{The effect strengthens on harder tasks.} The cross/same AT ratio increases from $7\times$ on CIFAR-10 to $26\times$ on CIFAR-100, while same-architecture divergence \emph{decreases} ($0.028 \to 0.010$), suggesting that more complex tasks drive architectures toward more specialized---and more divergent---representational strategies.
\end{enumerate}

These findings have immediate implications: representational insights from one architecture family may not transfer to another, and model ensembles combining CNNs with attention-based models may benefit precisely because they learn complementary features.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related Work}
\label{sec:related}

\textbf{Representation Similarity Metrics.} \citet{kornblith2019similarity} introduced CKA for comparing neural network representations, showing it is invariant to invertible linear transformations and orthogonal transformations. \citet{morcos2018insights} proposed SVCCA and PWCCA for subspace alignment analysis. These metrics measure how similarly networks organize data points in activation space but do not assess feature-level correspondence---whether individual neurons or channels encode the same information.

\textbf{Cross-Architecture Comparisons.} \citet{raghu2021vision} compared CNNs and ViTs using CKA, finding that ViTs develop more uniform representations across layers while CNNs preserve local spatial structure. \citet{nguyen2021wide} studied representation similarity within architecture families. Both studies rely on CKA, which---as we show---underestimates cross-architecture divergence.

\textbf{Feature Universality.} The hypothesis that different networks learn universal features \citep{li2015convergent,lenc2015understanding} has been influential but primarily tested within the same architecture family. Our work provides direct evidence against universality across architecture families.

\textbf{Optimal Transport in Deep Learning.} Optimal transport has been applied to domain adaptation \citep{courty2017joint}, model fusion \citep{singh2020model}, and generative modeling \citep{arjovsky2017wasserstein}. \citet{peyre2019computational} provide a comprehensive overview. We apply optimal transport to a new problem: quantifying feature-level correspondence across architectures.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Method}
\label{sec:method}

\subsection{Activation Transport Metric}

Given two trained networks $f$ and $g$ with feature dimensionalities $d_f$ and $d_g$, and a shared test set $\{x_i\}_{i=1}^N$, we extract penultimate-layer activations $F \in \mathbb{R}^{N \times d_f}$ and $G \in \mathbb{R}^{N \times d_g}$.

\textbf{Step 1: Standardization.} We z-score each channel independently:
\begin{equation}
    \hat{F}_{:,j} = \frac{F_{:,j} - \mu(F_{:,j})}{\sigma(F_{:,j}) + \epsilon}, \quad \epsilon = 10^{-8}
\end{equation}
This ensures AT measures distributional \emph{shape} similarity rather than scale differences.

\textbf{Step 2: Channel histograms.} For each channel, we compute a normalized histogram over $B = 25$ bins spanning $[-3, 3]$:
\begin{equation}
    h_j^F = \text{Hist}(\hat{F}_{:,j}; B), \quad h_j^F \leftarrow \frac{h_j^F + \epsilon}{\sum_k h_{j,k}^F + B\epsilon}
\end{equation}
Each histogram is a discrete probability distribution capturing the activation pattern of a single channel across the dataset.

\textbf{Step 3: Pairwise channel distances.} We compute the 1-Wasserstein distance between all pairs of channel histograms:
\begin{equation}
    C_{ij} = W_1(h_i^F, h_j^G) = \sum_{k=1}^{B-1} \left| \text{CDF}_i^F(k) - \text{CDF}_j^G(k) \right| \cdot \Delta
\end{equation}
where $\Delta = 6/B$ is the bin width and $\text{CDF}$ denotes the cumulative distribution function. This cost matrix $C \in \mathbb{R}^{d_f \times d_g}$ captures how dissimilar each pair of channels is.

\textbf{Step 4: Optimal assignment.} When $d_f \neq d_g$, we pad $C$ to a square matrix with penalty entries $10 \times \max(C)$ and solve the linear sum assignment problem \citep{kuhn1955hungarian}:
\begin{equation}
    \text{AT}(F, G) = \frac{1}{|\mathcal{V}|} \sum_{(i,j) \in \pi^*} C_{ij} \cdot \mathbf{1}[(i,j) \in \mathcal{V}]
\end{equation}
where $\pi^*$ is the optimal assignment from the Hungarian algorithm and $\mathcal{V}$ denotes valid (non-padded) pairs.

\textbf{Interpretation.} AT $\approx 0$ means every channel in $f$ has a distributional match in $g$---the networks learn interchangeable features. Large AT means the channel-wise activation patterns are fundamentally different, with no good one-to-one correspondence.

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=0.85, every node/.style={font=\small}]
  % Network A
  \draw[thick,rounded corners,fill=blue!10] (-0.5,3.5) rectangle (2.5,5.5);
  \node[font=\small\bfseries] at (1,5.2) {Network $f$};
  \node[font=\tiny] at (1,4.8) {(e.g., ResNet-18)};
  % Channels
  \foreach \y/\c in {4.3/ch$_1$, 4.0/ch$_2$, 3.7/ch$_3$} {
    \draw[fill=blue!30] (0.2,\y-0.1) rectangle (1.8,\y+0.1);
    \node[font=\tiny] at (1,\y) {\c};
  }

  % Network B
  \draw[thick,rounded corners,fill=red!10] (5.5,3.5) rectangle (8.5,5.5);
  \node[font=\small\bfseries] at (7,5.2) {Network $g$};
  \node[font=\tiny] at (7,4.8) {(e.g., ViT-Small)};
  \foreach \y/\c in {4.3/ch$_1$, 4.0/ch$_2$, 3.7/ch$_3$} {
    \draw[fill=red!30] (6.2,\y-0.1) rectangle (7.8,\y+0.1);
    \node[font=\tiny] at (7,\y) {\c};
  }

  % Step 1: Histograms
  \draw[->,thick] (1,3.4) -- (1,2.6);
  \draw[->,thick] (7,3.4) -- (7,2.6);
  \node[font=\tiny,anchor=east] at (0.8,3.0) {z-score};
  \node[font=\tiny,anchor=east] at (0.8,2.8) {+ hist};

  % Histogram boxes
  \draw[thick,rounded corners,fill=blue!5] (-0.3,1.2) rectangle (2.3,2.5);
  \node[font=\tiny] at (1,2.3) {Channel histograms};
  \draw[thick,rounded corners,fill=red!5] (5.7,1.2) rectangle (8.3,2.5);
  \node[font=\tiny] at (7,2.3) {Channel histograms};

  % Step 2: Cost matrix
  \draw[->,thick] (2.3,1.85) -- (3.2,1.85);
  \draw[->,thick] (5.7,1.85) -- (4.8,1.85);
  \draw[thick,rounded corners,fill=gray!15] (3.2,1.0) rectangle (4.8,2.7);
  \node[font=\tiny\bfseries] at (4.0,2.5) {Cost $C$};
  \node[font=\tiny] at (4.0,2.1) {$W_1(h_i^f, h_j^g)$};
  \node[font=\tiny] at (4.0,1.7) {for all};
  \node[font=\tiny] at (4.0,1.3) {$(i,j)$ pairs};

  % Step 3: Assignment
  \draw[->,thick] (4.0,0.9) -- (4.0,0.1);
  \node[font=\tiny,anchor=west] at (4.1,0.5) {Hungarian};

  % AT score
  \draw[thick,rounded corners,fill=green!15] (2.5,-0.6) rectangle (5.5,0.1);
  \node[font=\small\bfseries] at (4.0,-0.25) {AT $= \frac{1}{|\mathcal{V}|}\sum C_{ij}^*$};
\end{tikzpicture}
\caption{Overview of the Activation Transport pipeline. Given two trained networks, we (1)~extract and z-score penultimate-layer activations, (2)~compute per-channel histograms, (3)~build a pairwise cost matrix using the 1-Wasserstein distance between all channel histogram pairs, and (4)~solve the optimal assignment to find the lowest-cost channel matching. The average matched cost is the AT score.}
\label{fig:method}
\end{figure}

\subsection{Centered Kernel Alignment (Baseline)}

For comparison, we compute linear CKA \citep{kornblith2019similarity}:
\begin{equation}
    \text{CKA}(F, G) = \frac{\|F^T G\|_F^2}{\|F^T F\|_F \cdot \|G^T G\|_F}
\end{equation}
where $F$ and $G$ are mean-centered. CKA $\approx 1$ indicates similar representations. CKA measures global geometric alignment---how similarly networks organize the data manifold---but does not require or measure feature-level correspondence.

\subsection{Architectures}

We evaluate three architecturally distinct families:

\begin{enumerate}
    \item \textbf{ResNet-18.} A convolutional architecture with residual connections, learning hierarchical local features via spatial convolution. Final representation dimensionality: 512.
    \item \textbf{ViT-Small.} A vision transformer with global self-attention. For CIFAR-10: patch size 4, embedding dimension 192, 6 layers, 6 heads (final dim: 192). For CIFAR-100: patch size 4, embedding dimension 256, 8 layers, 8 heads (final dim: 256). The CIFAR-100 configuration uses a larger model because the $10\times$ increase in class count demands higher representational capacity from architectures that lack convolutional inductive biases. Without this scaling, ViT-Small fails to reach adequate accuracy on CIFAR-100 (below $40\%$), making AT comparisons meaningless. The ResNet-18 architecture is unchanged across datasets because its strong locality prior provides sufficient capacity for both settings. Note that this asymmetry in required scaling is itself evidence of the different representational strategies: CNNs generalize their local features efficiently to more classes, while attention-based models require more parameters.
    \item \textbf{MLP-Mixer.} A purely MLP-based architecture with token-mixing and channel-mixing layers. For CIFAR-10: patch size 4, dimension 192, 6 layers (final dim: 192). For CIFAR-100: patch size 4, dimension 256, 8 layers (final dim: 256), scaled identically to ViT to ensure the attention-family comparison remains controlled.
    \item \textbf{ResNet-18-v2 (same-architecture baseline).} An independently initialized ResNet-18, providing the same-architecture AT baseline.
\end{enumerate}

All models are trained with Adam (lr $= 10^{-3}$, weight decay $10^{-4}$), cosine annealing, batch size 128, and standard augmentation (random crop with padding 4, horizontal flip). For CIFAR-10: 50 max epochs, early stopping with patience 8, train/validation/test split 45K/5K/10K. For CIFAR-100: 80 max epochs, patience 10, same data split with 100 classes. CIFAR-10 experiments use 10 seeds (42--51); CIFAR-100 experiments use 10 seeds (42--51).

\subsection{Layer-Wise Analysis}

To study how divergence evolves through the network, we compute AT between intermediate representations at corresponding depth fractions. For ResNet-18, we extract features after layers 1--4 and the final global average pool. For ViT-Small, we extract the CLS token after transformer blocks 0, 2, 4, 5, and the final layer norm. These pairs represent approximately matched depth fractions (early, early-mid, mid-late, late, final).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
\label{sec:results}

We report results on CIFAR-10 aggregated over 10 independent seeds (42--51) and on CIFAR-100 aggregated over 10 seeds (42--51), with per-seed results in Appendix~\ref{app:seeds}.

\subsection{CIFAR-10: Cross-Architecture Divergence}

Table~\ref{tab:primary} presents the core results: pairwise AT and CKA scores between all architecture pairs on CIFAR-10.

\begin{table}[h]
\caption{Pairwise AT and CKA scores on CIFAR-10 (mean $\pm$ std, $n{=}10$ seeds).}
\label{tab:primary}
\centering
\small
\begin{tabular}{@{}lccc@{}}
\toprule
Architecture Pair & AT ($\downarrow$) & CKA ($\uparrow$) & Category \\
\midrule
ResNet-18 vs.\ ResNet-18-v2 & $0.028 \pm 0.004$ & $0.902 \pm 0.009$ & Same-arch \\
\midrule
ResNet-18 vs.\ ViT-Small    & $0.250 \pm 0.007$ & $0.633 \pm 0.017$ & CNN vs.\ Attention \\
ResNet-18 vs.\ MLP-Mixer    & $0.251 \pm 0.056$ & $0.660 \pm 0.017$ & CNN vs.\ Attention \\
ViT-Small vs.\ MLP-Mixer    & $0.088 \pm 0.035$ & $0.753 \pm 0.015$ & Attention vs.\ Attention \\
\bottomrule
\end{tabular}
\end{table}

The results reveal a clear hierarchy:

\begin{enumerate}
    \item \textbf{Same-architecture baseline.} Two ResNet-18s with different random initializations achieve AT $= 0.028$, indicating strong feature correspondence. CKA correspondingly reports $0.902$.
    \item \textbf{CNN vs.\ attention-based architectures.} ResNet-18 vs.\ ViT (AT $= 0.250$) and ResNet-18 vs.\ Mixer (AT $= 0.251$) show ${\sim}9\times$ higher divergence than same-architecture. These are fundamentally different representations.
    \item \textbf{Within attention-based family.} ViT vs.\ Mixer (AT $= 0.088$) is substantially lower than CNN-vs-attention pairs, suggesting attention-based and MLP-mixing architectures converge on more similar features---a shared ``attention family'' of representations distinct from CNNs.
\end{enumerate}

\textbf{CKA underestimates divergence.} While AT detects a $9\times$ gap between same-architecture and cross-architecture pairs, CKA reports only a modest decrease: $0.902 \to 0.633$ for ResNet-vs-ViT ($1.4\times$). CKA's global geometric measure---how similarly networks organize data points relative to each other---misses the feature-level divergence that AT captures. Two networks can organize data similarly (high CKA) while using completely different features to do so.

\begin{figure}[h]
\centering
\begin{tikzpicture}
  % AT scores (grouped bar chart placeholder)
  \draw[thick] (0,0) -- (0,3.5) -- (7,3.5) -- (7,0) -- cycle;
  \node[anchor=south] at (3.5,3.5) {\small\textbf{AT and CKA: Cross- vs.\ Same-Architecture}};
  % Bars representing AT scores (heights = value * 3.5 to fill 0-1 axis)
  \draw[fill=blue!70] (0.5,0) rectangle (1.1,0.28);   % same-arch AT 0.028
  \draw[fill=red!70]  (1.2,0) rectangle (1.8,2.50);    % ResNet-ViT AT 0.250
  \draw[fill=red!50]  (1.9,0) rectangle (2.5,2.51);    % ResNet-Mixer AT 0.251
  \draw[fill=orange!60](2.6,0) rectangle (3.2,0.88);   % ViT-Mixer AT 0.088
  % CKA bars
  \draw[fill=blue!30] (4.0,0) rectangle (4.6,3.16);    % same-arch CKA 0.902
  \draw[fill=red!30]  (4.7,0) rectangle (5.3,2.22);    % ResNet-ViT CKA 0.633
  \draw[fill=red!20]  (5.4,0) rectangle (6.0,2.31);    % ResNet-Mixer CKA 0.660
  \draw[fill=orange!20](6.1,0) rectangle (6.7,2.64);   % ViT-Mixer CKA 0.753
  % Y-axis labels
  \node[anchor=east,font=\tiny] at (0,0) {0.0};
  \node[anchor=east,font=\tiny] at (0,1.75) {0.5};
  \node[anchor=east,font=\tiny] at (0,3.5) {1.0};
  % X-axis labels
  \node[anchor=north,font=\tiny] at (1.85,0) {AT scores};
  \node[anchor=north,font=\tiny] at (5.35,0) {CKA scores};
  % Legend
  \draw[fill=blue!70] (0.3,3.1) rectangle (0.6,3.25); \node[anchor=west,font=\tiny] at (0.65,3.175) {Same-arch};
  \draw[fill=red!70]  (2.3,3.1) rectangle (2.6,3.25); \node[anchor=west,font=\tiny] at (2.65,3.175) {CNN--Attn};
  \draw[fill=orange!60](4.3,3.1) rectangle (4.6,3.25); \node[anchor=west,font=\tiny] at (4.65,3.175) {Attn--Attn};
\end{tikzpicture}
\caption{AT (left) reveals ${\sim}9\times$ divergence between same-architecture and cross-architecture pairs, while CKA (right) reports only ${\sim}1.4\times$ differences. Bar heights are proportional to measured values from Table~\ref{tab:primary}. This demonstrates that CKA's global geometric similarity measure misses the feature-level divergence that AT captures.}
\label{fig:at_vs_cka}
\end{figure}

\subsection{Accuracy Validation}

All models achieve competitive classification accuracy, confirming that AT differences reflect representational strategy rather than training quality:

\begin{table}[h]
\caption{Validation accuracy on CIFAR-10 (mean $\pm$ std, $n{=}10$ seeds).}
\label{tab:accuracy}
\centering
\small
\begin{tabular}{@{}lc@{}}
\toprule
Architecture & Val.\ Accuracy (\%) \\
\midrule
ResNet-18      & $93.27 \pm 0.17$ \\
ViT-Small      & $78.70 \pm 0.83$ \\
MLP-Mixer      & $81.49 \pm 0.53$ \\
\bottomrule
\end{tabular}
\end{table}

ResNet-18 achieves the highest accuracy ($93.3\%$), reflecting the strong inductive bias of convolutions for image classification at this scale. ViT-Small ($78.7\%$) and MLP-Mixer ($81.5\%$) achieve lower accuracy, which is expected: these architectures are designed for large-scale pretraining and lack the locality priors that make CNNs data-efficient on small images. Importantly, both attention-based models substantially exceed chance ($10\%$) and achieve accuracy comparable to early CNN architectures on CIFAR-10, confirming they learn meaningful visual features.

\textbf{Addressing the accuracy confound.} The $14.6$ percentage-point gap between ResNet and ViT raises a valid concern: does the high cross-architecture AT reflect genuine representational divergence or merely ViT's lower learning quality? Three observations argue against the accuracy-gap explanation:
\begin{enumerate}
    \item \textbf{Same-architecture AT is low despite identical accuracy.} ResNet-18 vs.\ ResNet-18-v2 yields AT $= 0.028$ despite both achieving $93.3\%$---showing that similar accuracy produces low AT, and thus high AT indicates representational, not quality, differences.
    \item \textbf{The ViT--Mixer comparison controls for accuracy.} ViT ($78.7\%$) and MLP-Mixer ($81.5\%$) have a similar accuracy gap ($2.8$ points) yet show low AT ($0.088$), demonstrating that accuracy differences per se do not drive high AT scores.
    \item \textbf{On CIFAR-100, the pattern strengthens.} Cross-architecture AT \emph{increases} from CIFAR-10 to CIFAR-100 ($0.250 \to 0.291$ for ResNet--ViT) while same-architecture AT \emph{decreases} ($0.028 \to 0.010$). If the accuracy gap were the primary driver, we would expect the signal to weaken or stay constant, not amplify.
\end{enumerate}
Nonetheless, a definitive resolution requires accuracy-controlled comparisons using pretrained models at matched accuracy levels. We discuss this further in Section~\ref{sec:limitations}.

\subsection{Layer-Wise Divergence}

Table~\ref{tab:layerwise} shows how AT between ResNet-18 and ViT-Small evolves across matched depth fractions.

\begin{table}[h]
\caption{Layer-wise AT between ResNet-18 and ViT-Small on CIFAR-10 (mean $\pm$ std, $n{=}10$ seeds).}
\label{tab:layerwise}
\centering
\small
\begin{tabular}{@{}llc@{}}
\toprule
ResNet Layer & ViT Block & AT \\
\midrule
Layer 1 (early)      & Block 0 & $0.034 \pm 0.004$ \\
Layer 2 (early-mid)  & Block 2 & $0.085 \pm 0.005$ \\
Layer 3 (mid-late)   & Block 4 & $0.184 \pm 0.006$ \\
Layer 4 (late)       & Block 5 & $0.255 \pm 0.007$ \\
Final (post-pool/norm) & Final  & $0.250 \pm 0.007$ \\
\bottomrule
\end{tabular}
\end{table}

Divergence increases \emph{monotonically} with depth: AT grows by $7.4\times$ from the earliest to latest layers ($0.034 \to 0.255$). This pattern has a natural interpretation: early layers in both architectures perform similar low-level processing (edge detection, texture extraction), regardless of whether this is implemented via convolution or attention. As depth increases, the architectures' different inductive biases lead to increasingly divergent representational strategies.

The slight decrease from layer 4 ($0.255$) to final ($0.250$) reflects the global average pooling in ResNet and layer normalization in ViT, which compress representations toward the classification interface.

\begin{figure}[h]
\centering
\begin{tikzpicture}[xscale=1.4, yscale=8]
  % Axes
  \draw[->] (0,0) -- (5.5,0) node[right,font=\small] {Depth};
  \draw[->] (0,0) -- (0,0.32) node[above,font=\small] {AT};
  % Y-axis ticks
  \foreach \y/\l in {0.0/0.00, 0.05/0.05, 0.10/0.10, 0.15/0.15, 0.20/0.20, 0.25/0.25} {
    \draw (-0.05,\y) -- (0.05,\y) node[left,font=\tiny,xshift=-2pt] {\l};
  }
  % X-axis ticks
  \foreach \x/\l in {1/Early, 2/Early-Mid, 3/Mid-Late, 4/Late, 5/Final} {
    \draw (\x,-0.005) -- (\x,0.005);
    \node[below,font=\tiny,rotate=20,anchor=north east] at (\x,-0.005) {\l};
  }
  % Data points and line (mean values from Table 3)
  \draw[thick,blue!70!black,mark=*] plot coordinates {(1,0.034) (2,0.085) (3,0.184) (4,0.255) (5,0.250)};
  % Error bars
  \foreach \x/\y/\e in {1/0.034/0.004, 2/0.085/0.005, 3/0.184/0.006, 4/0.255/0.007, 5/0.250/0.007} {
    \draw[blue!70!black] (\x,\y-\e) -- (\x,\y+\e);
    \draw[blue!70!black] (\x-0.05,\y-\e) -- (\x+0.05,\y-\e);
    \draw[blue!70!black] (\x-0.05,\y+\e) -- (\x+0.05,\y+\e);
    \fill[blue!70!black] (\x,\y) circle (0.4pt);
  }
  % Same-arch baseline
  \draw[dashed,gray] (0.5,0.028) -- (5.5,0.028);
  \node[right,font=\tiny,gray] at (5.5,0.028) {Same-arch};
\end{tikzpicture}
\caption{Layer-wise AT between ResNet-18 and ViT-Small on CIFAR-10 (mean $\pm$ std, $n{=}10$). Divergence increases monotonically with depth from $0.034$ (early) to $0.255$ (late), demonstrating that architectural inductive biases compound through the network. The dashed line shows the same-architecture baseline (AT $= 0.028$). Even the earliest layers already exceed this baseline, and by mid-depth the gap is $7\times$.}
\label{fig:layerwise}
\end{figure}

\subsection{CIFAR-100: Divergence Amplifies on Harder Tasks}
\label{sec:cifar100}

To test whether representational divergence is robust to increased task complexity, we repeat the experiment on CIFAR-100 (100 classes) with 10 independent seeds.

\begin{table}[h]
\caption{Pairwise AT and CKA scores on CIFAR-100 (mean $\pm$ std, $n{=}10$ seeds).}
\label{tab:cifar100}
\centering
\small
\begin{tabular}{@{}lccc@{}}
\toprule
Architecture Pair & AT ($\downarrow$) & CKA ($\uparrow$) & Category \\
\midrule
ResNet-18 vs.\ ResNet-18-v2 & $0.010 \pm 0.001$ & $0.745 \pm 0.007$ & Same-arch \\
\midrule
ResNet-18 vs.\ ViT-Small    & $0.291 \pm 0.003$ & $0.433 \pm 0.038$ & CNN vs.\ Attention \\
ResNet-18 vs.\ MLP-Mixer    & $0.356 \pm 0.091$ & $0.479 \pm 0.034$ & CNN vs.\ Attention \\
ViT-Small vs.\ MLP-Mixer    & $0.117 \pm 0.075$ & $0.528 \pm 0.048$ & Attention vs.\ Attention \\
\bottomrule
\end{tabular}
\end{table}

The CIFAR-100 results reveal a striking amplification of the cross-architecture divergence:

\begin{enumerate}
    \item \textbf{The cross/same ratio increases from $\mathbf{7\times}$ to $\mathbf{26\times}$.} Cross-architecture mean AT $= 0.254$ vs.\ same-architecture AT $= 0.010$ ($t = 11.00$, $p = 8.1 \times 10^{-16}$, two-sample $t$-test). The statistical significance is overwhelming.
    \item \textbf{Same-architecture divergence \emph{decreases}.} Two ResNet-18s trained on CIFAR-100 achieve AT $= 0.010$, compared to $0.028$ on CIFAR-10. More complex tasks appear to \emph{constrain} same-architecture representations toward a more unique solution, reducing the effect of random initialization.
    \item \textbf{Cross-architecture divergence increases.} ResNet vs.\ ViT AT rises from $0.250$ (CIFAR-10) to $0.291$ (CIFAR-100), and ResNet vs.\ Mixer from $0.251$ to $0.356$. Harder tasks drive architectures toward more specialized representations, amplifying the inter-family gap.
    \item \textbf{The attention family remains distinct.} ViT vs.\ Mixer AT ($0.117$) is still far below CNN-vs-attention pairs, confirming the attention family finding generalizes.
\end{enumerate}

\textbf{MLP-Mixer variance.} MLP-Mixer exhibits notably higher variance than other architectures across both datasets: ResNet vs.\ Mixer std $= 0.091$ on CIFAR-100 (vs.\ $0.003$ for ResNet vs.\ ViT), and ViT vs.\ Mixer std $= 0.075$ (Table~\ref{tab:per_seed_c100}). On CIFAR-10 (10 seeds), ResNet vs.\ Mixer shows std $= 0.056$ and ViT vs.\ Mixer shows std $= 0.035$, both substantially higher than the ResNet vs.\ ViT std of $0.007$. This instability likely reflects the MLP-Mixer's lack of strong inductive biases---without the locality of convolutions or the flexible contextualization of attention, the Mixer's learned features are more sensitive to random initialization. The Mixer's high variance does not affect our core finding (ResNet vs.\ ViT AT is remarkably tight: std $\leq 0.007$), but it suggests that the Mixer occupies a less stable region of representation space.

This scaling behavior has a natural interpretation: with 10 classes, there is slack in the representation---many feature configurations can solve the task, so different initializations and even different architectures partially overlap. With 100 classes, the representational demands are tighter: same-architecture networks converge on similar solutions (lower same-arch AT), while the distinct inductive biases of CNNs and attention models push them toward increasingly different feature spaces.

\subsection{Cross-Dataset Summary}

Table~\ref{tab:cross_dataset} summarizes the key metrics across both datasets.

\begin{table}[h]
\caption{Cross-dataset comparison of AT divergence metrics.}
\label{tab:cross_dataset}
\centering
\small
\begin{tabular}{@{}lcccc@{}}
\toprule
Dataset & Same-Arch AT & Cross-Arch AT & Ratio & $p$-value \\
\midrule
CIFAR-10 ($n{=}10$)  & $0.028$ & $0.197$ & $7.0\times$  & $3.8 \times 10^{-17}$ \\
CIFAR-100 ($n{=}10$) & $0.010$ & $0.254$ & $26.3\times$ & $8.1 \times 10^{-16}$ \\
\bottomrule
\end{tabular}
\end{table}

The $3.8\times$ increase in the cross/same ratio from CIFAR-10 to CIFAR-100 demonstrates that AT captures a fundamental architectural phenomenon that \emph{amplifies} with task complexity---precisely the opposite of what one would expect if architectural differences were superficial.

\subsection{ImageNet-Scale Validation}
\label{sec:imagenet}

To address the concern that our CIFAR results might reflect small-image artifacts and to eliminate the accuracy confound discussed in Section~\ref{sec:results}, we replicate the analysis at scale using publicly available \texttt{timm} checkpoints evaluated on ImageNet-V2 matched-frequency~\citep{recht2019imagenet}. ImageNet-V2 is the standard held-out test set for the post-2020 era and addresses train-test contamination concerns intrinsic to the original validation set. We sample 5{,}000 images per seed (random subsets of the 10K-image V2 pool) across three seeds (42, 43, 44).

\textbf{Models.} We evaluate three primary architectures and matched same-architecture variants:
\begin{itemize}
    \item \textbf{ResNet-50.} \texttt{resnet50} (timm A1 recipe, $69.13\%$ V2 top-1) and \texttt{resnet50.tv\_in1k} (torchvision recipe, $63.30\%$).
    \item \textbf{ViT-B/16.} \texttt{vit\_base\_patch16\_224} ($75.57\%$) and \texttt{deit\_base\_patch16\_224} (DeiT distillation, $71.22\%$).
    \item \textbf{MLP-Mixer-B/16.} \texttt{mixer\_b16\_224} ($62.97\%$) and \texttt{mixer\_b16\_224.miil\_in21k\_ft\_in1k} (MIIL ImageNet-21K-pretrained, $71.03\%$).
\end{itemize}

Each variant pair was verified to be a genuinely distinct checkpoint (mean weight differences in the range $0.04$--$0.27$). Each model is evaluated under the input preprocessing specified by its timm model card (per-model mean and standard deviation), following community benchmark conventions.

\textbf{Per-model preprocessing.} We let \texttt{timm} resolve each model's preferred normalization (\texttt{resolve\_model\_data\_config}) and feed each model the corresponding mean and standard deviation rather than reusing the same constants across models. All models receive $224 \times 224$ center-cropped inputs.

\textbf{ImageFolder label fix.} The \texttt{imagenetv2-matched-frequency-format-val} folder names directories by their \emph{integer ImageNet class index}, but PyTorch's \texttt{ImageFolder} alphabetizes the directory names and assigns labels in alphabetical order, which on this dataset corrupts the classification labels. We override \texttt{dataset.samples} so that the integer-named directory becomes the label, restoring the published ImageNet-V2 accuracy values.

\textbf{Feature extraction.} We register forward hooks at four matched depths per model: ResNet \texttt{layer1}--\texttt{layer4} for the convolutional family; transformer/mixer \texttt{blocks.\{2,5,8,11\}} for the attention/mixer families. Spatial / token dimensions are collapsed to per-sample channel vectors before computing histograms or kernel matrices. Each seed uses a different $5{,}000$-sample subset of the $10{,}000$-image ImageNet-V2 validation set, drawn under random seeds $\{42, 43, 44\}$ (the underlying pretrained model weights are fixed). On ImageNet-V2 our AT implementation is vectorized for speed and drops the bin-width prefactor $\Delta x$ from the cost in Section~\ref{sec:method}; this rescales ImageNet AT values uniformly relative to the CIFAR scale and does not affect the cross/same ratios reported below.

\textbf{Cross-architecture results.} Table~\ref{tab:imagenet_cross} reports AT and CKA between final-layer features.

\begin{table}[h]
\caption{Cross-architecture AT and CKA on ImageNet-V2 (mean $\pm$ std, $n{=}3$ seeds). Final-layer features.}
\label{tab:imagenet_cross}
\centering
\small
\begin{tabular}{@{}lcc@{}}
\toprule
Architecture Pair & AT ($\downarrow$) & CKA ($\uparrow$) \\
\midrule
ResNet-50 vs.\ ViT-B/16     & $2.9606 \pm 0.0006$ & $0.4791 \pm 0.0028$ \\
ResNet-50 vs.\ MLP-Mixer-B/16 & $2.6007 \pm 0.0007$ & $0.3165 \pm 0.0019$ \\
ViT-B/16 vs.\ MLP-Mixer-B/16  & $0.5028 \pm 0.0003$ & $0.4326 \pm 0.0052$ \\
\bottomrule
\end{tabular}
\end{table}

ViT-vs-Mixer AT ($0.50$) is $5.9\times$ smaller than ResNet-vs-ViT AT ($2.96$), replicating the ``attention family'' finding from CIFAR at full ImageNet scale.

\textbf{Same-architecture baselines.} Table~\ref{tab:imagenet_same} reports same-architecture AT for each family.

\begin{table}[h]
\caption{Same-architecture AT and CKA on ImageNet-V2 (mean $\pm$ std, $n{=}3$ seeds).}
\label{tab:imagenet_same}
\centering
\small
\begin{tabular}{@{}lccp{4cm}@{}}
\toprule
Pair & AT ($\downarrow$) & CKA ($\uparrow$) & Notes \\
\midrule
ViT-B/16 vs.\ DeiT-B & $\mathbf{0.0489 \pm 0.0004}$ & $0.7368 \pm 0.0016$ & Cleanest baseline \\
ResNet-50 vs.\ ResNet-50-tv & $2.1201 \pm 0.0011$ & $0.5033 \pm 0.0025$ & Different training recipe (timm A1 vs.\ torchvision) \\
Mixer-B/16 vs.\ Mixer-MIIL & $1.0517 \pm 0.0000$ & $0.4063 \pm 0.0017$ & MIIL is in21k-pretrained, not seed variation \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Methodology: anchoring the same-architecture baseline.} An unweighted average across all three same-architecture pairs yields a same-arch AT of $\sim 1.07$ and a cross/same ratio of $\sim 1.88\times$. However, two of the three pairs differ in training recipe rather than random seed (ResNet timm-A1 vs.\ torchvision; Mixer in1K vs.\ in21K-pretrained), so they are not interchangeable baselines. We therefore anchor headline ratios on the architecturally cleanest pair (ViT-DeiT) and report the full robustness grid below.

\textbf{Headline cross/same ratios.} Anchored on the ViT-DeiT clean baseline:

\begin{table}[h]
\caption{Cross/same AT ratios on ImageNet-V2, anchored on ViT-DeiT same-architecture baseline. Welch's two-sample $t$-test.}
\label{tab:imagenet_ratios}
\centering
\small
\begin{tabular}{@{}lccccc@{}}
\toprule
Cross-Architecture Pair & Cross AT & Same AT & Ratio & $t$ & $p$ \\
\midrule
ResNet vs.\ ViT     & $2.9606$ & $0.0489$ & $\mathbf{60.5\times}$ & $6597$ & $1.05 \times 10^{-13}$ \\
ResNet vs.\ Mixer   & $2.6007$ & $0.0489$ & $\mathbf{53.2\times}$ & $5398$ & $9.40 \times 10^{-13}$ \\
ViT vs.\ Mixer      & $0.5028$ & $0.0489$ & $\mathbf{10.3\times}$ & $1500$ & $2.40 \times 10^{-11}$ \\
\bottomrule
\end{tabular}
\end{table}

All three cross-architecture comparisons are massively significant ($p < 10^{-10}$) and yield ratios $\geq 10\times$ \emph{when anchored on the ViT-DeiT baseline}. We caution that this anchor is the architecturally cleanest of the three available baselines but is not the only defensible choice; we therefore report a robustness analysis below.

\textbf{Robustness to baseline choice.} Because the headline ratio depends on which same-architecture pair is taken as the reference, we report cross/same ratios for ResNet-vs-ViT computed against \emph{each} of the three same-architecture baselines, at every depth fraction. Table~\ref{tab:imagenet_robust} shows the full grid.

\begin{table}[h]
\caption{Robustness of the ResNet-vs-ViT cross/same AT ratio under each of three same-architecture baselines, computed at four depth fractions. Cross AT is the same in every column. Each entry is the cross AT divided by the corresponding same-architecture baseline AT \emph{at the matched layer index}. The ViT-DeiT baseline is architecturally cleanest; the other two are confounded by training-recipe (ResNet) or pretraining-data (Mixer) differences.}
\label{tab:imagenet_robust}
\centering
\small
\begin{tabular}{@{}lcccc@{}}
\toprule
Layer & Cross AT & ViT-DeiT base & ResNet recipe base & Mixer-MIIL base \\
\midrule
L1 (early)         & $1.467$ & $\mathbf{17.9\times}$ & $2.32\times$ & $2.36\times$ \\
L2 (mid-early)     & $1.767$ & $\mathbf{34.2\times}$ & $3.18\times$ & $1.19\times$ \\
L3 (mid-late)      & $2.146$ & $\mathbf{55.0\times}$ & $6.11\times$ & $1.51\times$ \\
L4 (final)         & $2.961$ & $\mathbf{60.5\times}$ & $1.40\times$ & $2.81\times$ \\
\bottomrule
\end{tabular}
\end{table}

The cross/same AT separation is robust in direction (\emph{all 12 cells} show ratio $> 1$), with a worst-case ratio of $1.19\times$ (L2, Mixer-MIIL baseline) and median $3.0\times$. The ViT-DeiT baseline is markedly the cleanest---increasing monotonically with depth and yielding the largest separations---and we accordingly anchor headline numbers there. Two patterns are notable in the confounded baselines: (i) the ResNet recipe baseline shoots up at L4 (AT $= 2.12$), explaining the $1.40\times$ ratio there, because the timm A1 vs.\ torchvision recipes diverge specifically in their final-classifier feature space rather than in their early visual processing; (ii) the Mixer-MIIL baseline is anomalously high at every layer because ImageNet-21K pretraining produces fundamentally different representations than ImageNet-1K-only training. Both anomalies are evidence \emph{against} treating these as clean ``same architecture, same training, different seed'' proxies.

\textbf{CKA contrast.} CKA again fails to track architectural divergence at the same magnitude. Same-arch CKA (ViT-DeiT $= 0.737$) is only $1.5\times$ the cross-arch CKA (ResNet-vs-ViT $= 0.479$). For AT, the ratio is $60\times$. This $40\times$ disparity in sensitivity between AT and CKA is the central empirical claim of the paper, now confirmed at ImageNet scale.

\textbf{Layer-wise depth scaling.} Computing AT at four depth fractions through each model, we observe:
\begin{itemize}
    \item \textbf{ResNet vs.\ ViT:} $1.467 \to 1.767 \to 2.146 \to 2.961$ (monotonically increasing).
    \item \textbf{ResNet vs.\ Mixer:} $1.399 \to 1.379 \to 1.688 \to 2.601$ (monotonically increasing).
    \item \textbf{ViT vs.\ Mixer:} $1.139 \to 0.748 \to 0.721 \to 0.503$ (monotonically \emph{decreasing}).
\end{itemize}
Both CNN-vs-attention pairs replicate the monotonic depth-amplification pattern observed on CIFAR. Notably, the ViT-vs-Mixer pair shows the opposite---representations \emph{converge} with depth---further supporting the ``attention family'' claim: rather than diverging through the network, attention-based architectures align more closely as features become more task-specific.

\textbf{Mixer same-architecture caveat.} The Mixer-vs-MIIL AT of $1.05$ is uncomfortably high for a putative ``same-architecture'' baseline. We attribute this to the fundamental difference between ImageNet-1K-only training and ImageNet-21K pretraining: the latter produces representations specialized for a $21\times$ larger label space, which differ substantially from 1K-only features even within the same architecture. We therefore do not anchor the headline ratio on this pair, and we report it only to characterize the dependence of same-architecture AT on training distribution.

\textbf{Cross-seed stability.} Standard deviations across the three seeds are on the order of $10^{-3}$, indicating that the metric saturates well before sample size becomes an issue---5{,}000 samples are sufficient for stable estimates at this scale.

\textbf{Summary of ImageNet findings.} (1)~Cross-architecture AT exceeds same-architecture AT under every baseline choice we considered, with a worst-case ratio of $1.19\times$ and a best-case ratio of $60.5\times$ (anchored on the architecturally cleanest ViT-DeiT baseline). (2)~The ``attention family'' clustering replicates: ViT and Mixer remain $5.9\times$ closer to each other than either is to ResNet. (3)~Monotonic depth amplification replicates for CNN-vs-attention pairs. (4)~CKA again fails to detect this divergence at comparable magnitude. The dependence of headline ratio on baseline choice is a real limitation: a true ``same-architecture, same-training, different-seed'' baseline at ImageNet scale would require training multiple instances from scratch ($\sim$2-3 GPU-days each), which we leave to future work. The current robustness analysis (Table~\ref{tab:imagenet_robust}) shows the qualitative direction of the AT separation is preserved across all three available proxies and all four depth fractions ($12/12$ cells positive).

\begin{figure}[h]
\centering
\begin{tikzpicture}[xscale=2.5, yscale=10]
  % Axes
  \draw[->] (0,0) -- (3.2,0) node[right,font=\small] {};
  \draw[->] (0,0) -- (0,0.30) node[above,font=\small] {AT};
  % Y-axis ticks
  \foreach \y/\l in {0.0/0.00, 0.05/0.05, 0.10/0.10, 0.15/0.15, 0.20/0.20, 0.25/0.25} {
    \draw (-0.02,\y) -- (0.02,\y) node[left,font=\tiny,xshift=-2pt] {\l};
  }
  % CIFAR-10 bars
  \draw[fill=blue!30] (0.3,0) rectangle (0.7,0.028);
  \draw[fill=red!60]  (0.8,0) rectangle (1.2,0.197);
  \node[below,font=\small] at (0.75,-0.01) {CIFAR-10};
  % CIFAR-100 bars
  \draw[fill=blue!30] (1.8,0) rectangle (2.2,0.010);
  \draw[fill=red!60]  (2.3,0) rectangle (2.7,0.254);
  \node[below,font=\small] at (2.25,-0.01) {CIFAR-100};
  % Ratio annotations
  \draw[<->,thick] (1.25,0.028) -- (1.25,0.197);
  \node[right,font=\tiny] at (1.25,0.112) {$7\times$};
  \draw[<->,thick] (2.75,0.010) -- (2.75,0.254);
  \node[right,font=\tiny] at (2.75,0.130) {$26\times$};
  % Legend
  \draw[fill=blue!30] (0.3,0.27) rectangle (0.5,0.28); \node[right,font=\tiny] at (0.52,0.275) {Same-arch};
  \draw[fill=red!60]  (1.5,0.27) rectangle (1.7,0.28); \node[right,font=\tiny] at (1.72,0.275) {Cross-arch};
\end{tikzpicture}
\caption{Cross-dataset comparison of AT divergence. The cross-architecture/same-architecture AT ratio increases from $7\times$ on CIFAR-10 to $26\times$ on CIFAR-100. Same-architecture AT \emph{decreases} from $0.028$ to $0.010$ as harder tasks constrain representations, while cross-architecture AT \emph{increases} from $0.197$ to $0.254$, demonstrating that task complexity amplifies architectural divergence.}
\label{fig:cross_dataset}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
\label{sec:discussion}

\subsection{Why CKA Misses the Divergence}

CKA measures whether two networks organize data points similarly in activation space---specifically, whether pairs of inputs that are close in one network's representation are also close in the other's. Two networks can have high CKA while using entirely different features, as long as both features lead to the same relative arrangement of data points.

Consider an analogy: two cities might have identical population density maps (high CKA) while being built with completely different construction materials and architectural styles (high AT). CKA measures the ``density map''; AT measures the ``building materials.''

This distinction matters because many applications depend on feature-level correspondence rather than global geometric similarity. Transfer learning, feature distillation, and mechanistic interpretability all require that specific neurons or channels encode specific information---exactly what AT measures.

\subsection{The Attention Family}

The low AT between ViT and MLP-Mixer ($0.088$ on CIFAR-10, $0.117$ on CIFAR-100) despite their very different computational mechanisms (quadratic self-attention vs.\ linear token mixing) suggests that the critical architectural choice is not \emph{how} tokens communicate but \emph{whether} communication is spatially local (convolution) or global (attention/mixing). Both ViT and Mixer process all spatial positions simultaneously, while CNNs build representations through local receptive field growth. This global-vs-local distinction appears to determine the ``family'' of learned representations more than the specific mixing mechanism.

\subsection{Implications}

\textbf{Transfer learning.} Feature-level divergence between CNNs and transformers suggests that naive feature transfer across architecture families will be ineffective. Intermediate representations from a CNN may not provide useful initialization for a transformer, and vice versa.

\textbf{Model ensembling.} The high cross-architecture AT scores suggest that CNN-transformer ensembles may benefit from genuine representational diversity rather than merely noise averaging. This provides a feature-level explanation for why mixed-architecture ensembles often outperform same-architecture ensembles.

\textbf{Mechanistic interpretability.} Interpretability insights---such as identifying specific neurons as ``edge detectors'' or ``texture detectors''---may be architecture-specific. The monotonically increasing divergence with depth suggests that early-layer insights are more likely to transfer across architectures than late-layer insights.

\textbf{Architecture search.} AT provides a principled metric for measuring representational diversity in neural architecture search, complementing accuracy-based evaluation with a measure of \emph{how differently} an architecture processes information.

\subsection{Depth-Dependent Divergence}

The monotonic increase in AT with depth ($0.034 \to 0.255$) has a natural information-theoretic interpretation. Early layers in any image classifier must extract basic visual features (edges, textures, colors), and there are limited ways to do this regardless of architecture. As depth increases, each architecture applies its inductive bias repeatedly: convolutions build increasingly large receptive fields through local composition, while attention layers maintain global context from the start. These fundamentally different computational strategies lead to increasingly divergent representational states.

This finding connects to \citet{raghu2021vision}'s observation that ViTs develop more spatially uniform representations than CNNs. Our AT analysis quantifies this divergence at the feature level and shows it is not merely a difference in spatial structure but a fundamental difference in what information individual channels encode.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Limitations}
\label{sec:limitations}

\begin{enumerate}
    \item \textbf{Scale.} Our primary experiments use CIFAR-10 and CIFAR-100 ($32 \times 32$). Section~\ref{sec:imagenet} replicates the core findings on ImageNet-V2 using base-scale pretrained models (ResNet-50, ViT-B/16, Mixer-B/16); extending to larger backbones (e.g., ResNet-152, ViT-L/16) and to additional pretraining-data regimes is a natural next step.
    \item \textbf{Seeds.} We use 10 seeds on both CIFAR-10 and CIFAR-100. The tight confidence intervals (e.g., ResNet vs.\ ViT AT std $= 0.007$ on CIFAR-10, $0.003$ on CIFAR-100) suggest the results are stable, but additional seeds would further strengthen the statistical claims. On ImageNet-V2, the three ``seeds'' are random subsets of the same evaluation pool rather than independently trained model instances; a true seed-level ImageNet baseline would require training multiple instances from scratch ($\sim$2--3 GPU-days each), which we leave to future work.
    \item \textbf{Histogram granularity.} The 25-bin histogram discretization may lose fine-grained distributional information. Alternative approaches using kernel density estimation or exact Wasserstein computation could provide smoother estimates.
    \item \textbf{Layer matching.} Matching ResNet layers to ViT blocks by depth fraction is approximate. Alternative alignment strategies (e.g., by computational cost or representational capacity) might reveal different patterns.
    \item \textbf{Feature granularity.} AT operates at the channel level. Sub-channel features (e.g., polysemantic neurons) would require finer-grained analysis.
    \item \textbf{Accuracy gap (CIFAR only).} On CIFAR, ViT ($78.7\%$) and Mixer ($81.5\%$) achieve substantially lower accuracy than ResNet ($93.3\%$). The ImageNet-V2 replication in Section~\ref{sec:imagenet} addresses this concern directly: ResNet-50 ($69.13\%$), ViT-B/16 ($75.57\%$), and Mixer-B/16 ($62.97\%$) span comparable accuracy ranges, and the cross/same AT separation is even larger ($60.5\times$) than on CIFAR-100 ($26\times$). This argues strongly against the accuracy gap as the primary driver of high cross-architecture AT.
    \item \textbf{ViT/Mixer architecture scaling.} On CIFAR-100, we scale ViT and Mixer from (dim=192, depth=6) to (dim=256, depth=8) while keeping ResNet-18 fixed. Although this is necessary for adequate ViT/Mixer performance, it introduces a confound: cross-architecture AT comparisons on CIFAR-100 involve different-capacity models. Future work should test whether the divergence persists when all architectures are exactly matched in parameter count.
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
\label{sec:conclusion}

We introduce Activation Transport (AT), an optimal transport-based metric for measuring feature-level correspondence between neural network architectures. Across three datasets, AT reveals that cross-architecture feature divergence between CNNs and attention-based models is an order of magnitude larger than same-architecture divergence: $7\times$ on CIFAR-10 ($n{=}10$; $t = 31.27$, $p = 3.8 \times 10^{-17}$), $26\times$ on CIFAR-100 ($n{=}10$; $t = 11.00$, $p = 8.1 \times 10^{-16}$), and $60.5\times$ on ImageNet-V2 using base-scale pretrained models ($n{=}3$ seeds; $t = 6597$, $p = 1.05 \times 10^{-13}$). CKA dramatically underestimates this gap, reporting only $1.4\times$--$1.5\times$ differences for the same comparisons across all three datasets.

The divergence amplifies with task complexity: on CIFAR-100, same-architecture AT \emph{decreases} ($0.028 \to 0.010$) while cross-architecture AT \emph{increases} ($0.197 \to 0.254$), demonstrating that harder tasks drive architectures toward more specialized and divergent feature spaces. Layer-wise analysis shows divergence increases monotonically with depth ($0.034 \to 0.255$), confirming that architectural inductive biases compound rather than converge. Within the attention-based family, ViT and MLP-Mixer show consistently lower divergence (AT $\approx 0.10$), indicating a shared representational strategy distinct from CNNs.

These findings challenge the assumption that networks converge to universal representations when trained on the same task and establish AT as a more sensitive tool than CKA for detecting representation-level architectural differences. Our results have direct implications for transfer learning, model ensembling, and the scope of mechanistic interpretability.

\textbf{Future directions.} Two natural extensions of this work are (1)~\emph{training dynamics}: measuring AT between ResNet and ViT at checkpoints throughout training to determine when divergence emerges (e.g., immediately, or only after early layers have converged), and (2)~\emph{normalization effects}: testing whether the choice of normalization layer (BatchNorm, LayerNorm, GroupNorm) affects cross-architecture AT, which would reveal whether normalization acts as a ``feature attractor'' that shapes representations independently of the core architecture. Pilot experiment code for both directions exists in our repository.

Code and data are available at \url{https://anonymous.4open.science/r/activation-transport}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{plainnat}
\bibliography{refs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\appendix
\section{Per-Seed Results: CIFAR-10}
\label{app:seeds}

\begin{table}[h]
\caption{Per-seed AT and CKA scores on CIFAR-10 (seeds 42--46 shown; full $n{=}10$ aggregates in Table~\ref{tab:primary}).}
\label{tab:per_seed_c10}
\centering
\small
\begin{tabular}{@{}lcccccc@{}}
\toprule
 & \multicolumn{2}{c}{ResNet vs.\ ViT} & \multicolumn{2}{c}{ResNet vs.\ Mixer} & \multicolumn{2}{c}{ResNet vs.\ ResNet2} \\
Seed & AT & CKA & AT & CKA & AT & CKA \\
\midrule
42 & 0.264 & 0.638 & 0.247 & 0.669 & 0.023 & 0.901 \\
43 & 0.252 & 0.646 & 0.227 & 0.680 & 0.025 & 0.906 \\
44 & 0.259 & 0.629 & 0.225 & 0.662 & 0.026 & 0.906 \\
45 & 0.247 & 0.661 & 0.125 & 0.671 & 0.024 & 0.906 \\
46 & 0.250 & 0.652 & 0.220 & 0.668 & 0.027 & 0.905 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\caption{Per-seed layer-wise AT between ResNet-18 and ViT-Small on CIFAR-10 (seeds 42--46 shown; full $n{=}10$ aggregates in Table~\ref{tab:layerwise}).}
\label{tab:per_seed_layer}
\centering
\small
\begin{tabular}{@{}ccccc@{}}
\toprule
Seed & Layer 1 & Layer 2 & Layer 3 & Layer 4 \\
\midrule
42 & 0.036 & 0.089 & 0.186 & 0.269 \\
43 & 0.039 & 0.081 & 0.180 & 0.257 \\
44 & 0.042 & 0.092 & 0.183 & 0.262 \\
45 & 0.050 & 0.085 & 0.176 & 0.250 \\
46 & 0.030 & 0.093 & 0.181 & 0.253 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\caption{Per-seed validation accuracies on CIFAR-10 (seeds 42--46 shown; full $n{=}10$ aggregates in Table~\ref{tab:accuracy}).}
\label{tab:per_seed_acc}
\centering
\small
\begin{tabular}{@{}ccccc@{}}
\toprule
Seed & ResNet-18 & ViT-Small & MLP-Mixer & ResNet-18-v2 \\
\midrule
42 & 93.34 & 79.68 & 81.18 & 93.40 \\
43 & 93.02 & 77.72 & 81.64 & 93.28 \\
44 & 93.02 & 77.10 & 82.08 & 93.16 \\
45 & 93.22 & 78.60 & 80.58 & 93.30 \\
46 & 93.36 & 79.22 & 81.16 & 93.04 \\
\bottomrule
\end{tabular}
\end{table}

\section{Per-Seed Results: CIFAR-100}
\label{app:seeds_c100}

\begin{table*}[t]
\caption{Per-seed AT scores on CIFAR-100 ($n{=}10$ seeds). Note the remarkably tight standard deviation for ResNet vs.\ ViT (std $= 0.003$) and ResNet vs.\ ResNet2 (std $= 0.001$).}
\label{tab:per_seed_c100}
\centering
\small
\begin{tabular}{@{}ccccc@{}}
\toprule
Seed & ResNet vs.\ ViT & ResNet vs.\ Mixer & ViT vs.\ Mixer & ResNet vs.\ ResNet2 \\
\midrule
42 & 0.293 & 0.209 & 0.110 & 0.011 \\
43 & 0.295 & 0.421 & 0.160 & 0.011 \\
44 & 0.292 & 0.330 & 0.058 & 0.008 \\
45 & 0.286 & 0.289 & 0.018 & 0.009 \\
46 & 0.286 & 0.412 & 0.161 & 0.012 \\
47 & 0.292 & 0.349 & 0.070 & 0.009 \\
48 & 0.292 & 0.382 & 0.121 & 0.008 \\
49 & 0.290 & 0.363 & 0.093 & 0.010 \\
50 & 0.293 & 0.319 & 0.044 & 0.009 \\
51 & 0.293 & 0.506 & 0.268 & 0.009 \\
\bottomrule
\end{tabular}
\end{table*}

\section{Per-seed ImageNet AT (3 seeds $\times$ 6 model pairs $\times$ 4 layers)}
\label{app:seeds_imagenet}

For the ImageNet-V2 experiments in Section~\ref{sec:imagenet}, the three ``seeds'' are random $5{,}000$-image subsets of the same ImageNet-V2 evaluation pool drawn under random seeds $\{42, 43, 44\}$ (the underlying pretrained model weights are fixed). Table~\ref{tab:per_seed_imagenet_final} reports the final-layer AT for each of the six model pairs, and Table~\ref{tab:per_seed_imagenet_layers} reports the layer-wise AT at four matched depth fractions (L1--L4) for each pair. Standard deviations across the three seeds are on the order of $10^{-3}$, indicating that the metric saturates well before sample size becomes an issue at this scale.

\begin{table}[h]
\caption{Per-seed final-layer AT on ImageNet-V2 for the three same-architecture and three cross-architecture pairs ($n{=}3$ seeds; full aggregates in Tables~\ref{tab:imagenet_same}--\ref{tab:imagenet_ratios}).}
\label{tab:per_seed_imagenet_final}
\centering
\small
\begin{tabular}{@{}lccc@{}}
\toprule
Pair & Seed 42 & Seed 43 & Seed 44 \\
\midrule
\multicolumn{4}{l}{\emph{Same-architecture}} \\
ResNet-50 vs.\ ResNet-50-tv             & $2.1214$ & $2.1193$ & $2.1195$ \\
ViT-B/16 vs.\ DeiT-B                    & $0.0489$ & $0.0485$ & $0.0493$ \\
Mixer-B/16 vs.\ Mixer-MIIL              & $1.0517$ & $1.0517$ & $1.0518$ \\
\midrule
\multicolumn{4}{l}{\emph{Cross-architecture}} \\
ResNet-50 vs.\ ViT-B/16                 & $2.9612$ & $2.9600$ & $2.9605$ \\
ResNet-50 vs.\ Mixer-B/16               & $2.6014$ & $2.6000$ & $2.6008$ \\
ViT-B/16 vs.\ Mixer-B/16                & $0.5026$ & $0.5031$ & $0.5026$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\caption{Per-seed layer-wise AT on ImageNet-V2 at four matched depth fractions (L1--L4) for each of the six pairs. For ResNet-based pairs the ResNet stages \texttt{layer1}--\texttt{layer4} are matched to ViT/Mixer blocks \texttt{2, 5, 8, 11}; for ViT/Mixer-only pairs we use blocks \texttt{2, 5, 8, 11} on both sides.}
\label{tab:per_seed_imagenet_layers}
\centering
\small
\begin{tabular}{@{}llcccc@{}}
\toprule
Pair & Seed & L1 & L2 & L3 & L4 \\
\midrule
\multirow{3}{*}{ResNet-50 vs.\ ResNet-50-tv}
 & 42 & $0.6348$ & $0.5608$ & $0.3516$ & $2.1214$ \\
 & 43 & $0.6318$ & $0.5518$ & $0.3509$ & $2.1193$ \\
 & 44 & $0.6329$ & $0.5545$ & $0.3513$ & $2.1195$ \\
\midrule
\multirow{3}{*}{ViT-B/16 vs.\ DeiT-B}
 & 42 & $0.0817$ & $0.0517$ & $0.0394$ & $0.0489$ \\
 & 43 & $0.0817$ & $0.0517$ & $0.0388$ & $0.0485$ \\
 & 44 & $0.0822$ & $0.0518$ & $0.0389$ & $0.0493$ \\
\midrule
\multirow{3}{*}{Mixer-B/16 vs.\ Mixer-MIIL}
 & 42 & $0.6215$ & $1.4787$ & $1.4234$ & $1.0517$ \\
 & 43 & $0.6213$ & $1.4785$ & $1.4234$ & $1.0517$ \\
 & 44 & $0.6213$ & $1.4786$ & $1.4234$ & $1.0518$ \\
\midrule
\multirow{3}{*}{ResNet-50 vs.\ ViT-B/16}
 & 42 & $1.4676$ & $1.7663$ & $2.1461$ & $2.9612$ \\
 & 43 & $1.4675$ & $1.7673$ & $2.1465$ & $2.9600$ \\
 & 44 & $1.4660$ & $1.7663$ & $2.1450$ & $2.9605$ \\
\midrule
\multirow{3}{*}{ResNet-50 vs.\ Mixer-B/16}
 & 42 & $1.3989$ & $1.3794$ & $1.6879$ & $2.6014$ \\
 & 43 & $1.3994$ & $1.3785$ & $1.6881$ & $2.6000$ \\
 & 44 & $1.3977$ & $1.3782$ & $1.6869$ & $2.6008$ \\
\midrule
\multirow{3}{*}{ViT-B/16 vs.\ Mixer-B/16}
 & 42 & $1.1393$ & $0.7475$ & $0.7210$ & $0.5026$ \\
 & 43 & $1.1377$ & $0.7476$ & $0.7210$ & $0.5031$ \\
 & 44 & $1.1390$ & $0.7479$ & $0.7211$ & $0.5026$ \\
\bottomrule
\end{tabular}
\end{table}

\section{Experimental Configuration}
\label{app:config}

\begin{table}[h]
\caption{Full experimental configuration.}
\label{tab:config}
\centering
\small
\begin{tabular}{@{}lp{10cm}@{}}
\toprule
Parameter & Value \\
\midrule
\textbf{ResNet-18} & BasicBlock, $[2,2,2,2]$, channels $[64,128,256,512]$ \\
\textbf{ViT-Small (CIFAR-10)} & Patch size 4, dim 192, depth 6, heads 6, MLP dim 384 \\
\textbf{ViT-Small (CIFAR-100)} & Patch size 4, dim 256, depth 8, heads 8, MLP dim 512 \\
\textbf{MLP-Mixer (CIFAR-10)} & Patch size 4, dim 192, depth 6, token MLP dim 96, channel MLP dim 384 \\
\textbf{MLP-Mixer (CIFAR-100)} & Patch size 4, dim 256, depth 8, token MLP dim 128, channel MLP dim 512 \\
Datasets & CIFAR-10 (45K/5K/10K), CIFAR-100 (45K/5K/10K) \\
Optimizer & Adam (lr $= 10^{-3}$, weight decay $= 10^{-4}$) \\
Scheduler & Cosine annealing ($T_{\max} = $ max epochs) \\
Epochs & CIFAR-10: 50 (early stopping, patience 8); CIFAR-100: 80 (patience 10) \\
Batch size & 128 \\
Augmentation & RandomCrop(32, pad=4), RandomHorizontalFlip \\
AT bins & 25, range $[-3, 3]$ \\
Seeds & 42--51 (CIFAR-10), 42--51 (CIFAR-100) \\
Device & NVIDIA A100 SXM 80GB \\
\bottomrule
\end{tabular}
\end{table}

\end{document}
